---
title: "Homework 1 - Solutions"
format: 
  html:
    toc: true
    theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = T, cache = T)
```


1.  (2 pts) Load the rankinSAV.csv dataset as a dataframe or tibble. Short answer question 1: What are the dimension of the dataset?

```{r}
library(tidyverse)

df = read_csv('data/rankinSAV.csv')
dim(df)

```
Short answer: `r nrow(df)` rows and `r ncol(df)` columns

2.  (3 pts) Add 2 new columns to the dataframe/tibble. One should contain the sum of the cover of all of the seagrass species (total_seagrass), and the other should contain the the sum of the cover of all of the macroalgae species (total_macroalgae). Short answer question 2: What is the maximum value for total seagrass and total macroalgae and in what years did they occur?

```{r}
df = read_csv('data/rankinSAV.csv') |> 
  mutate(total_seagrass = TT + HW + SF,
         total_macroalgae = HAL + BAT + PEN) 

# total seagrass
ts = df |> 
  filter(total_seagrass == max(total_seagrass, na.rm =T))

ts

# total seagrass
tm = df |> 
  filter(total_macroalgae == max(total_macroalgae, na.rm =T))

tm
```

Short answer: The max total seagrass was `r ts$total_seagrass` in `r ts$YEAR`. The max total macroalgae was `r tm$total_macroalgae` in `r tm$YEAR`.

3.  (2 pts) Create a new tibble/dataframe that contains only the year, total seagrass, and total macroalgae.

```{r}
d = df |> 
  select(YEAR, total_seagrass, total_macroalgae)
```


4.  (5 pts) Convert the new tibble/dataframe from question 3 into long format for the two SAV metrics and summarize the mean and SD for each taxonomic group in each year. Short answer question 3: What are the years with the highest and lowest cover for seagrass and macroalgae?

```{r}
dm = d |> 
  pivot_longer(total_seagrass:total_macroalgae, 
               names_to = 'SAV', values_to = 'cover') |> 
  group_by(YEAR, SAV) |> 
  summarize(mean = mean(cover, na.rm = T),
            sd = sd(cover, na.rm = T))

dm

sg = dm |> filter(SAV == 'total_seagrass')

sg$YEAR[which.max(sg$mean)]
sg$YEAR[which.min(sg$mean)]

ma = dm |> filter(SAV == 'total_macroalgae')

ma$YEAR[which.max(ma$mean)]
ma$YEAR[which.min(ma$mean)]
```

Short answer: The year with the highest seagrass cover is `r sg$YEAR[which.max(sg$mean)]` and lowest is `r sg$YEAR[which.min(sg$mean)]`. The year with the highest macroalgae cover is `r ma$YEAR[which.max(ma$mean)]` and lowest is `r ma$YEAR[which.min(ma$mean)]`.

5.  (3 pts) Plot the mean cover for each year for all 6 species of SAV, with a different species in a different panel.
```{r}
lg = df |> 
  group_by(YEAR) |> 
  summarize(across(TT:PEN,mean) )|> 
  pivot_longer(TT:PEN, names_to = 'SAV', values_to = 'cover')

ggplot(lg, aes(YEAR, cover, color = SAV)) +
  geom_point() +
  geom_line(size = 1) +
  labs(x = 'Time', y = 'SAV cover', color = 'Species') +
  facet_wrap(~SAV)+
  theme_bw()
```


6.  (2 pts) Convert the percent cover of seagrass into three categories: low = \< 0.30, intermediate = 0.31-0.60, and high = \> 0.60.

```{r}
df = df |> 
  mutate(sg = case_when(
    total_seagrass <= 0.30 ~ 'low',
    between(total_seagrass, 0.30,0.60) ~ 'intermediate',
    total_seagrass > 0.60 ~ 'high'
  ))
```


7.  (5 pts) Create a conditional (i.e. `TRUE`/`FALSE`) vector to compare if seagrass cover is higher compared to the year before. ***Hint** - this can be done with a for loop*

```{r}

# for loop
s = df |> 
  group_by(YEAR) |> 
  summarize(sg = mean(total_seagrass)) |> 
  mutate(comp = NA)
for (i in 2:nrow(s)){
  s$comp[i] = s$sg[i] > s$sg[i-1]
}
  

#dplyr
sg = df |> 
  group_by(YEAR) |> 
  summarize(sg = mean(total_seagrass)) |> 
  mutate(comp = sg>lag(sg))



```


8.  (3 pts) Plot the count of years where seagrass cover is higher than the year before and the count of years where seagrass cover is lower than the year before.
```{r}
sgc = s |> 
  count(comp) |> 
  drop_na()

ggplot(sgc, aes(comp, n, fill = comp)) +
  geom_col() +
  labs(x = 'SG cover(t) > sG cover(t-1)', y = 'count') +
  theme_bw()+
  theme(legend.position = 'none')
```

